---
title: "BIOST 546 HW 1"
author: "My-Anh Doan"
date: "2023-01-09"
output:
  pdf_document: default
editor_options:
  chunk_output_type: console
---

```{r chunk-setup, include = FALSE}
# set global options for code chunks
knitr::opts_chunk$set(message = FALSE, warning = FALSE, collapse = TRUE)
```

Q1. In this problem, we will make use of the dataset `Medical_Cost_2.RData`.

(a) Load the dataset with the command `load` and check if there are missing data.  
(b) If any, remove the missing data using the command `na.omit`.
```{r load-data}
# load data to environment
load("Medical_Cost_2.RData") # dataframe with 1338 rows and 7 variables

# check for missing data; if returned value is non-zero, there are missing data
sum(is.na(df) == TRUE) # 60

# remove missing data
df <- na.omit(df) # dataframe with 1278 rows and 7 variables
```

(c) We decide to focus on the outcome variable `charges` (individual medical costs billed by health insurance) and the predictors `bmi` (body mass index) and `smoker` (whether the subjects is a smoker or not). Make a scatterplot with `bmi` on the x-axis, `charges` on the y-axis, and with the color of each dot representing whether the subject is a smoker or not.
```{r fig-1}
library(ggplot2)

ggplot(df, aes(x = bmi, y = charges)) +
  geom_point(aes(color = smoker)) +
  labs(x = "BMI",
       y = "Charges",
       color = "Smoker status") +
  theme_bw() +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_blank())

```

